Applying WaveNet on frequency data

### 2017-01-20 paper/paper.tex
Revied section "Introduction" and "Models and Methods"

\section{Introduction}
In 2016, \cite{wavenet} introduced WaveNet, a generative model for raw audio achieving astonishing results in learning sound features and generating smooth audio. Trained on piano music, new highly dynamic piano music varying in melody, volume and speed smoothly could be generated.\\
The aim this paper is to enhance the WaveNet architecture and apply it to a more feature-oriented representation of input data. Instead of performing a classification task on raw audio, the network is adopted to operate on the frequencies of the audio data, processed by Fouier transformation. In particular, the following topics are covered:
\begin{itemize}
\item Based on WaveNet from \cite{wavenet} a network architecture to generate music is introduced operating on the frequencies of audio input.
\item This new network is trained and evaluated on piano music from the MusicNet dataset. Furthermore, optimization approaches as data normalization and variations in preprocessing the data are included in the experiments.
\item Results from the evaluation are analyzed in detail. In the context of the introduced network, advantages as well as weaknesses are discussed. 
\end{itemize}
As training data in the experiments, piano samples from the MusicNet dataset, a collection of 330 freely-licenced recordings of classical music are taken. The data is publicly available online \cite{musicnet}.\\
The source code of the experiments can be found on GitHub.\footnote{\url{https://github.com/sennluzi/tensorflow-wavenet}}

\section{Models and Methods}
The following section describes the underlying model of the network used in the experiments of this paper. As core network the WaveNet architecture with a stack of dilated causal convolutional layers (figure \ref{fig:wavenet_dilated_cnn}) is taken. Hereupon, the data input layer is changed conceptually.
\begin{figure}[tbp]
  \centering
  \includegraphics[width=\columnwidth]{figures/wavenet_dilated_cnn.png}
  \caption{dilated convolutional network}
  \label{fig:wavenet_dilated_cnn}
\end{figure}

\subsection{WaveNet architecture}
Given a sequence of amplitudes, WaveNet predicts the subsequent amplitude. The amplitudes are quantized in 256 bins \cite{itu1988711}. In the $i$th step, the network then outputs a probability distribution $P(x_{i+1} | x_1, ..., x_i)$. The amplitude $x_{i+1}$ of the next step is then sampled according to that distribution.\\
In WaveNet, training can be done in parallel while the generation of a audio sample is done sequentially.\\
As a basis for WaveNet, the implementation of Igor Babuschkin is taken \cite{Babuschkin2016}.\\
The original WaveNet paper mentiones only very few coefficients for how their network is configured. Therefore reproducing the paper exactly is very hard. The following numbers are assumptions mainly based on the experience of Babuschkin's implementation. In the following experiments, the dilations are $2^n$ for $n$ taking the values $(0,1,...,7,8)$ and are repeated 5 times. This leads to 50 hidden dilation layers. The convolution filter size is set to $2$ and the number of channels of the residual and the dilation is $32$.\\
The input of the WaveNet network is a sequence of amplitudes, each of which encoded as a one-hot vector.

\textcolor{red}{rough description of wavenet architecture and specification used during the experiments}
\subsection{Data preprocessing}
To extract frequencies, the raw audio data consisting of pressure measurements are divided into chunks of size 2048, which are then transformed into the frequency space using FFT. Therefrom, the first 150 coefficients are taken as dominant frequencies for the subsequent training. To ensure that audio features are transformed smoothly, two neighboring raw audio chunks overlap in a few samples, described by a variable $stride$. Equation \ref{eq:fft} summarized the frequency extraction of the $i^\textnormal{th}$ chunk compactly.
\begin{multline}\label{eq:fft}
\textnormal{frequencies\_coeffs}_i =
FFT(\textnormal{inputData}[i \cdot stride:\\i \cdot stride+2048])[:150]
\end{multline}
Considering real and imaginary parts of the complex Fourier coefficients separately, each chunk of 2048 raw audio samples is transformed into 300 real frequency values.\\
To decrease the dimensionality even further, principle component analysis (PCA) is used to reduce these 300 frequencies to 100 values.
These 100 PCA coefficient are considered one tone and fed into the network as one time step.\\
The eigenvectors of the PCA are shown in figure \ref{fig:pca_eigenvectors_imaginary_numbers}. Figure \ref{fig:pca_eigenvectors} illustrates how the PCA looks like when taking the absolute values of the imaginary numbers, out of which the components are built from. In this figure, one can see, that the relevant frequencies are all relatively low. Figure \ref{fig:pca_explained_variance} shows that a vast majority of the variance of the frequencies is captured by the first 100 PCA components. Summarizing, in neither of the two steps, we loose much of the original input information. This should lead to a decent compression of the audio signal.\\
To make this evident, one can figure invert all the steps and plot the compressed wave and the original wave (see \ref{fig:preprocessing_difference_waveform}). It is easy to see that they resemble a lot. After preprocessing the audio contains a cracking noise, but nevertheless is the melody and the tone of the music well audible.\\
To ensure this preprocessing still preserves the nature of the music, one can invert all steps. Although after preprocessing, there is an audible cracking, the music then is still clearly hearable and the pitches recognizable.
On one hand, the waveform looks still similar (figure \ref{fig:preprocessing_difference_waveform})
Whereas in the original WaveNet architecture, 16000 samples represented on second of audio, the adopted network only needs to process $16000/2048 \approx 8$ samples per second. However, this change infers a fundamentally different approach of audio training and generation, described in the next section.

>>>>

\section{Introduction}
In 2016, \cite{wavenet} introduced WaveNet, a generative model for raw audio achieving astonishing results in learning sound features and generating harmonious audio. Trained on piano music, highly dynamic piano music varying smoothly in melody, volume and speed could be generated.\\
The aim of this paper is to enhance the WaveNet architecture and apply it to a more feature-oriented representation of input data. Instead of performing a classification task on raw audio, the network is adopted to operate on the frequencies of the audio data, processed by Fouier transformation.\\
As training data in the experiments, piano samples from the MusicNet dataset, a collection of 330 freely-licenced recordings of classical music are taken. The data is publicly available online \cite{musicnet}.\\
In particular, the following topics are covered:
\begin{itemize}
\item Based on WaveNet from \cite{wavenet}, a network architecture to generate music is introduced operating on the frequencies of audio input.
\item This new network is trained and evaluated on piano music from the MusicNet dataset. Furthermore, optimization approaches as data normalization and variations in preprocessing the data are included in the experiments.
\item Results from the evaluation are analyzed in detail. In the context of the introduced network, advantages as well as weaknesses are discussed. 
\end{itemize}
The source code of the experiments can be found on GitHub.\footnote{\url{https://github.com/sennluzi/tensorflow-wavenet}}

\section{Models and Methods}
The following section describes the underlying model of the network used in the experiments of this paper. As core network the WaveNet architecture with a stack of dilated causal convolutional layers is taken. Hereupon, the data input layer is changed conceptually.
\begin{figure}[tbp]
	\centering
	\includegraphics[width=\columnwidth]{figures/wavenet_dilated_cnn.png}
	\caption{dilated convolutional network}
	\label{fig:wavenet_dilated_cnn}
\end{figure}
\subsection{WaveNet architecture}
Given a sequence of amplitudes, WaveNet predicts the subsequent amplitude. This task is done in a classification setting: The amplitudes are quantized in 256 classes \cite{itu1988711}. The input of the WaveNet network is a sequence of such quantizied amplitudes, each encoded as a 256-dimensional one-hot vector. Performing a softmax-function in its last layer, the network outputs a probability distribution $P(x_{i+1} | x_1, ..., x_i)$ in the $i$th step. The amplitude $x_{i+1}$ of the next step is then sampled according to that distribution.\\
To model the the dependency of an audio sample on previous samples, WaveNet uses a stack of dilated convolutions. Figure \ref{fig:wavenet_dilated_cnn} depicts this architecture exemplarily for such a network of depth 4. By using exponentially growing dilations in depth, a network with $n$ dilated convolutional layers can observe the last $2^n$ input values to predict an output.\\
Training WaveNet is done similar as an autoencoder: The $i$th amplitude of the input data is predicted by only using previous input data and taken as label for the loss calculation and backpropagation. Thus, training only requires the input data and can be done in parallel. Against that, the generation of a audio sample has to be processed sequentially, since the $i$th step relies on the output of the $(i-1)$th one.\\
As a basis for WaveNet, the implementation of Igor Babuschkin is taken \cite{Babuschkin2016}. The original WaveNet paper mentions only very few coefficients for how their network is configured. Therefore reproducing the paper exactly is very hard. The following numbers are assumptions mainly based on the experience of Babuschkin's implementation. In the following experiments, the dilations are $2^n$ for $n$ taking the values $(0,1,...,8)$ and are repeated 5 times. This leads to 50 hidden dilation layers. The convolution filter size is set to $2$ and the number of channels of the residual and the dilation is $32$.\\
\subsection{Data preprocessing}
To extract frequencies, the raw audio data consisting of pressure measurements are divided into chunks of size 2048, which are then transformed into the frequency space using FFT. Therefrom, the first 150 coefficients are taken as dominant frequencies for the subsequent training. To ensure that audio features are transformed smoothly, two neighboring raw audio chunks overlap in a few samples, described by a variable $stride$. Equation \ref{eq:fft} summarized the frequency extraction of the $i^\textnormal{th}$ chunk compactly.
\begin{multline}\label{eq:fft}
\textnormal{frequencies\_coeffs}_i =
FFT(\textnormal{inputData}[i \cdot stride:\\i \cdot stride+2048])[:150]
\end{multline}
Considering real and imaginary parts of the complex Fourier coefficients separately, each chunk of 2048 raw audio samples is transformed into 300 real frequency values.\\
To decrease the dimensionality even further, principle component analysis (PCA) is used to reduce these 300 frequencies to 100 values.
These 100 PCA coefficient are considered one tone and fed into the network as one time step.\\
The eigenvectors of the PCA are shown in figure \ref{fig:pca_eigenvectors_imaginary_numbers}. Figure \ref{fig:pca_eigenvectors} illustrates how the PCA looks like when taking the absolute values of the imaginary numbers, out of which the components are built from. In this figure, one can see, that the relevant frequencies are all relatively low. Figure \ref{fig:pca_explained_variance} shows that a vast majority of the variance of the frequencies is captured by the first 100 PCA components. Summarizing, in neither of the two steps, we loose much of the original input information. This should lead to a decent compression of the audio signal.\\
To make this evident, one can figure invert all the steps and plot the compressed wave and the original wave (see \ref{fig:preprocessing_difference_waveform}). It is easy to see that they resemble a lot. After preprocessing the audio contains a cracking noise, but nevertheless is the melody and the tone of the music well audible.\\
To ensure this preprocessing still preserves the nature of the music, one can invert all steps. Although after preprocessing, there is an audible cracking, the music then is still clearly hearable and the pitches recognizable.
On one hand, the waveform looks still similar (figure \ref{fig:preprocessing_difference_waveform})
Whereas in the original WaveNet architecture, 16000 samples represented on second of audio, the adopted network only needs to process $16000/2048 \approx 8$ samples per second. However, this change infers a fundamentally different approach of audio training and generation, described in the next section.
